---
title: "Simulation"
author: "Thu Dang"
format: html
---


# Warming Up

Most statistical software packages have several built-in distributions to handle the most common situations statisticians come across when simulating values.

Remember to set your seeds for these exercises so that you get the same answer each time the code is evaluated. 

## Exponential Distribution

1. Simulate 1000 draws from the exponential distribution $$f(x) = \lambda e^{-\lambda x},\;\;\;\; x\geq0, \text{ with } \lambda=3$$

```{r}
# Question 1
# set a default seed
set.seed(402)

lambda <- 3 
x <- rexp(1000, rate = lambda)

head(x)
```

2. Create a density plot of your values. Add the PDF of the exponential(3) distribution to your plot in a different color. How do they compare?

```{r}
# Question 2
# loading package
library(ggplot2)

df <- data.frame(x)
ggplot(df, aes(x)) +
    geom_density(linewidth = 1.0, aes(color = "Simulated")) + 
    stat_function(fun = function(x) lambda * exp(-lambda * x), aes(color = "true pdf"), linewidth = 1.0) +
    xlab("x") + ylab("Density") +
    ggtitle("Density of simulated data + true p.d.f") +
    labs(color = "Legend")
```

----

> The simulated density curve closely matches with the shape of the exponential(3) p.d.f.


3. How might you use your 1000 draws to estimate the probability that $x > 1$? Provide an estimate using only your 1000 draws, and compare this estimate to one derived from `pexp` (R) or `scipy.stats.expon.cdf` (Python).

```{r}
# Question 3
# estimate probability of simulation
p_simulation <- sum(x > 1) / length(x)

# true probability
p_true <- 1 - pexp(1, rate = lambda)

p_simulation
p_true
```

----

> I estimated `P(x>1)` using 1000 simulated values by calculate the proportion of draws greater than 1. This gave an estimate of about 0.047. But using the true exponential distribution, the result is 0.04978707. The simulation estimate is very close to the true value.

4. What is the average error from using a $n=1000$ simulation instead of a numerical calculation for the probability that $x>1$? Write a function to calculate the error from one simulation, and evaluate that function 100 times to calculate the expected error from using a simulation estimate rather than deterministic numerical integration.

```{r}
# Question 4
# creating a function
error_simulation <- function(n = 1000, lambda = 3) {
  x <- rexp(n, rate = lambda)
  p_simulation <- sum(x > 1) / length(x)
  p_true <- 1 - pexp(1, rate = lambda)
  abs(p_simulation - p_true)
}

# run 100 simulations
set.seed(402)
errors <- replicate(100, error_simulation())
mean(errors)
```
----

> This shows that using a simulation with n = 1000 gives a very accurate estimate of `P(X>1)`.



## Gamma Distribution

The Gamma distribution is an exponential family, so it has MLEs that are unbiased and efficient estimators, if somewhat challenging to calculate in closed form. 

Luckily, there are functions in R and python to handle this:

- `fitdistr(...)` in the `MASS` package (R) will estimate the  parameters for a gamma distribution from sample data.

- `scipy.stats` distribution functions have a `.fit(data)` method that will estimate distributional parameters

In this problem, you will work with the gamma distribution with parameters $\alpha, \beta$ (note that R uses a different parameterization than Python) and PDF 

$$f(x | \alpha, \beta) = \frac{\beta^\alpha x^{\alpha-1} e^{-x}}{\Gamma(\alpha)}.$$

1. Write a function that will calculate the MLE $\hat\alpha,\hat\beta$ from a vector $x_{samp}$ of gamma-distributed samples. Your function should have required parameter $x$ and optional parameters $\alpha,\beta$ representing the true values. The function should return a list of $\hat\alpha, \hat\beta$ and, if $\alpha,\beta$ are provided, it should also return $\alpha_{err} = \alpha-\hat\alpha$ and $\beta_{err} = \beta-\hat\beta$. 

```{r}
# Question 1
library(MASS)

fit_gamma <- function(x, alpha = NULL, beta = NULL) {
  fit <- fitdistr(x, "gamma")
  # extract estimates
  alpha_hat <- fit$estimate["shape"]
  beta_hat  <- fit$estimate["rate"]
  # create output list
  output <- list(alpha_hat = alpha_hat, beta_hat = beta_hat)
  # if providing true alpha or beta -> calculate errors
  if (!is.null(alpha)) output$alpha_err <- alpha - alpha_hat 
  if (!is.null(beta)) output$beta_err  <- beta - beta_hat  
  
  return(output)
}

set.seed(403)
x_sample <- rgamma(1000, shape = 2, rate = 3)
fit_gamma(x_sample, alpha = 2, beta = 3)
```

2. For a grid of $\alpha \times \beta=\{0.25, 0.50, 1.00, 2.00, 4.00\}\times\{0.25, 0.50, 1.00, 2.00, 4.00\}$, sample 100 values from the $\text{gamma}(\alpha,\beta)$ distribution. Provide the true parameter values and plot your error values in an appropriate plot. Take care to choose your plot mappings to support your discussion of the following: What trends (if any) do you notice in the estimation error?

```{r}
# Question 2
# loading package
library(dplyr)
library(ggplot2)

# creating a 5x5 grid
alpha_value <- c(0.25, 0.50, 1.00, 2.00, 4.00)
beta_value  <- c(0.25, 0.50, 1.00, 2.00, 4.00)
grid <- expand.grid(alpha = alpha_value, beta = beta_value)

set.seed(403)
results <- data.frame()

for (i in 1:nrow(grid)) {
  a <- grid$alpha[i]
  b <- grid$beta[i]
  # draw 100 values
  x_sample <- rgamma(100, shape = a, rate = b)
  # estimate
  estimate <- fit_gamma(x_sample, alpha = a, beta = b)
 
  results <- rbind(results, data.frame(alpha = a, beta = b,                                                                        alpha_err = estimate$alpha_err, beta_err = estimate$beta_err))
}
# plot for alpha
ggplot(results, aes(x = alpha, y = beta, fill = alpha_err)) +
  geom_tile() +
  scale_fill_gradient2(mid = "lightblue") +
  xlab("true alpha") + ylab("true beta") +
  ggtitle("Alpha estimation error") +
  labs(fill = "alpha_err")

# plot for beta
ggplot(results, aes(x = alpha, y = beta, fill = beta_err)) +
  geom_tile() +
  scale_fill_gradient2(mid = "white") +
  xlab("true alpha") + ylab("true beta") +
  ggtitle("Beta estimation error") +
  labs(fill = "beta_err")
```

----

> From the two plots, the estimation errors might not follow a very clear pattern. Because each cell is based on only one random sample of size 100. I can see that some larger alpha and beta give errors closer to 0, but there are also cases where the error isn't small.The results show that the errors might vary a lot for different parameter when the sample size is only 100.


3. Now, let's vary the sample size. For $n=\{20, 30, 50, 100\}$, use your function to evaluate the error for $\alpha=1, \beta=2$. How does the error change as $n$ increases? Provide an appropriate plot as well as a description of the changes for increasing $n$. 


```{r}
# Question 3
n_values <- c(20, 30, 50, 100)

set.seed(403)

results_n <- data.frame()

for (n in n_values) {
  x_sample <- rgamma(n, shape = 1, rate = 2)
  
  estimate_n <- fit_gamma(x_sample, alpha = 1, beta = 2)
  
  results_n <- rbind(results_n,data.frame(n = n, alpha_err = estimate_n$alpha_err,                                                                              beta_err  =  estimate_n$beta_err))
}

results_n

# plot for alpha
ggplot(results_n, aes(x = n, y = alpha_err)) +
  geom_point(size = 3, color = "blue") +
  geom_line(color = "blue") +
  xlab("Sample size") + ylab("Alpha error") +
  ggtitle("Alpha error vs sample size") 

# plot for beta
ggplot(results_n, aes(x = n, y = beta_err)) +
  geom_point(size = 3, color = "darkred") +
  geom_line(color = "darkred") +
  xlab("Sample size") + ylab("Beta error") +
  ggtitle("Beta error vs sample size") 
```

----

> The errors change a lot when the sample size is small. For n=20 and n=30, alpha and beta errors might be quite large. When the sample size increases to 50 and 100, the errors become smaller and closer to zero.
This shows that the MLE becomes more stable when having more data.
