---
title: "Simulation"
author: "Thu Dang"
format: html
---


# Warming Up

Most statistical software packages have several built-in distributions to handle the most common situations statisticians come across when simulating values.

Remember to set your seeds for these exercises so that you get the same answer each time the code is evaluated. 

## Exponential Distribution

1. Simulate 1000 draws from the exponential distribution $$f(x) = \lambda e^{-\lambda x},\;\;\;\; x\geq0, \text{ with } \lambda=3$$


2. Create a density plot of your values. Add the PDF of the exponential(3) distribution to your plot in a different color. How do they compare?

3. How might you use your 1000 draws to estimate the probability that $x > 1$? Provide an estimate using only your 1000 draws, and compare this estimate to one derived from `pexp` (R) or `scipy.stats.expon.cdf` (Python).

4. What is the average error from using a $n=1000$ simulation instead of a numerical calculation for the probability that $x>1$? Write a function to calculate the error from one simulation, and evaluate that function 100 times to calculate the expected error from using a simulation estimate rather than deterministic numerical integration.

```{r}
# Question 1
# set a default seed
set.seed(402)

lambda <- 3 
x <- rexp(1000, rate = lambda)

head(x)
```

```{r}
# Question 2
# loading package
library(ggplot2)

df <- data.frame(x)
ggplot(df, aes(x)) +
    geom_density(linewidth = 1.0, aes(color = "Simulated")) + 
    stat_function(fun = function(x) lambda * exp(-lambda * x), aes(color = "true pdf"), linewidth = 1.0) +
    xlab("x") + ylab("Density") +
    ggtitle("Density of simulated data + true p.d.f") +
    labs(color = "Legend")
```

----

> The simulated density curve closely matches with the shape of the exponential(3) p.d.f.


```{r}
# Question 3
# estimate probability of simulation
p_simulation <- sum(x > 1) / length(x)

# true probability
p_true <- 1 - pexp(1, rate = lambda)

p_simulation
p_true
```

----

> I estimated `P(x > 1)` using 1000 simulated values by calculate the proportion of draws greater than 1. This gave an estimate of about 0.047. But using the true exponential distribution, the result is 0.04978707. The simulation estimate is very close to the true value.


```{r}
# Question 4
# creating a function
error_simulation <- function(n = 1000, lambda = 3) {
  x <- rexp(n, rate = lambda)
  p_simulation <- sum(x > 1) / length(x)
  p_true <- 1 - pexp(1, rate = lambda)
  abs(p_simulation - p_true)
}

# run 100 simulations
set.seed(402)
errors <- replicate(100, error_simulation())
mean(errors)
```

## Gamma Distribution

The Gamma distribution is an exponential family, so it has MLEs that are unbiased and efficient estimators, if somewhat challenging to calculate in closed form. 

Luckily, there are functions in R and python to handle this:

- `fitdistr(...)` in the `MASS` package (R) will estimate the  parameters for a gamma distribution from sample data.

- `scipy.stats` distribution functions have a `.fit(data)` method that will estimate distributional parameters

In this problem, you will work with the gamma distribution with parameters $\alpha, \beta$ (note that R uses a different parameterization than Python) and PDF 

$$f(x | \alpha, \beta) = \frac{\beta^\alpha x^{\alpha-1} e^{-x}}{\Gamma(\alpha)}.$$

1. Write a function that will calculate the MLE $\hat\alpha,\hat\beta$ from a vector $x_{samp}$ of gamma-distributed samples. Your function should have required parameter $x$ and optional parameters $\alpha,\beta$ representing the true values. The function should return a list of $\hat\alpha, \hat\beta$ and, if $\alpha,\beta$ are provided, it should also return $\alpha_{err} = \alpha-\hat\alpha$ and $\beta_{err} = \beta-\hat\beta$. 

```{r}
# Question 1
library(MASS)

fit_gamma <- function(x, alpha = NULL, beta = NULL) {
  fit <- fitdistr(x, densfun = "gamma")
  # extract estimates
  alpha_hat <- fit$estimate["shape"]
  beta_hat  <- fit$estimate["rate"]
  # create output list
  output <- list(alpha_hat = alpha_hat, beta_hat = beta_hat)
  # if true alpha or beta provided -> errors
  if (!is.null(alpha) && !is.null(beta)) {
    output$alpha_err <- alpha - alpha_hat
    output$beta_err  <- beta - beta_hat
  }
  
  return(output)
}

set.seed(402)
x_sample <- rgamma(1000, shape = 2, rate = 3)
fit_gamma(x_sample, alpha = 2, beta = 3)
```

2. For a grid of $\alpha \times \beta=\{0.25, 0.50, 1.00, 2.00, 4.00\}\times\{0.25, 0.50, 1.00, 2.00, 4.00\}$, sample 100 values from the $\text{gamma}(\alpha,\beta)$ distribution. Provide the true parameter values and plot your error values in an appropriate plot. Take care to choose your plot mappings to support your discussion of the following: What trends (if any) do you notice in the estimation error?

```{r}
# Question 2
# loading package
library(dplyr)
library(ggplot2)

alpha_grid <- c(0.25, 0.5, 1, 2, 4)
beta_grid  <- c(0.25, 0.5, 1, 2, 4)
grid <- expand.grid(alpha = alpha_grid, beta = beta_grid)

set.seed(402)
results <- data.frame()

for (i in 1:nrow(grid)) {
  a <- grid$alpha[i]
  b <- grid$beta[i]
  # sample 100 values
  x_sample <- rgamma(100, shape = a, rate = b)
  # estimate
  estimate <- fit_gamma(x_sample, alpha = a, beta = b)
 
  results <- rbind(results, data.frame(alpha = a, beta = b,                                                                        alpha_err = estimate$alpha_err, beta_err = estimate$beta_err))
}
# plot for alpha
ggplot(results, aes(x = alpha, y = beta, fill = alpha_err)) +
  geom_tile() +
  scale_fill_gradient2(mid = "white") +
  xlab("true alpha") + ylab("true beta") +
  ggtitle("Alpha estimation error") +
  labs(fill = "alpha_err")

# plot for beta
ggplot(results, aes(x = alpha, y = beta, fill = beta_err)) +
  geom_tile() +
  scale_fill_gradient2(mid = "white") +
  xlab("true alpha") + ylab("true beta") +
  ggtitle("Beta estimation error") +
  labs(fill = "beta_err")
```

----

> The estimation error is largest when `alpha` and `beta` are small. When the parameters are large, the errors shrink significantly. This makes sense because small alpha produces highly skewed and noisy samples, which makes MLE estimation less stable, especially with only 100 observations. In both heatmaps, the error patterns show that: Error decreases as `alpha` increases. Error decreases as `beta` increases.


3. Now, let's vary the sample size. For $n=\{20, 30, 50, 100\}$, use your function to evaluate the error for $\alpha=1, \beta=2$. How does the error change as $n$ increases? Provide an appropriate plot as well as a description of the changes for increasing $n$. 


```{r}
# Question 3

alpha_true <- 1
beta_true  <- 2

n_values <- c(20, 30, 50, 100)

set.seed(402)

results_n <- data.frame()

for (n in n_values) {
  x_sample <- rgamma(n, shape = alpha_true, rate = beta_true)
  
  est <- fit_gamma(x_sample, alpha = alpha_true, beta = beta_true)
  
  results_n <- rbind(results_n,data.frame(n = n, alpha_err = est$alpha_err,beta_err  = est$beta_err))
}

results_n

# plot for alpha
ggplot(results_n, aes(x = n, y = alpha_err)) +
  geom_point(size = 3, color = "blue") +
  geom_line(color = "blue") +
  xlab("Sample size") + ylab("Alpha error") +
  ggtitle("Alpha error vs sample size") 

# plot for beta
ggplot(results_n, aes(x = n, y = beta_err)) +
  geom_point(size = 3, color = "darkred") +
  geom_line(color = "darkred") +
  xlab("Sample size") + ylab("Beta error") +
  ggtitle("Beta error vs sample size") 
```

----

> As the sample size increases, the estimation error decreases. When n = 20, both alpha and beta errors are relatively large and unstable. The errors become smaller at n = 30 and n = 50, and by the time n reaches 100, the estimates are very close to the true parameters.
